---
title: "Data With Danny: Data Bank"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```


```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "casestudy3", 
                      Trusted_Connection = "True")
```

### Task 

There is a new innovation in the financial industry called Neo-Banks: new aged digital only banks without physical branches.

Danny thought that there should be some sort of intersection between these new age banks, cryptocurrency and the data world…so he decides to launch a new initiative - Data Bank!

Data Bank runs just like any other digital bank - but it isn’t only for banking activities, they also have the world’s most secure distributed data storage platform!

Customers are allocated cloud data storage limits which are directly linked to how much money they have in their accounts. There are a few interesting caveats that go with this business model, and this is where the Data Bank team need your help!

The management team at Data Bank want to increase their total customer base - but also need some help tracking just how much data storage their customers will need.

This case study is all about calculating metrics, growth and helping the business analyse their data in a smart way to better forecast and plan for their future developments!

### Avaiable Data
The Data Bank team have prepared a data model for this case study as well as a few example rows from the complete dataset below to get you familiar with their tables.

#### Entity Relationship Diagram

#### Table 1: Regions
Just like popular cryptocurrency platforms - Data Bank is also run off a network of nodes where both money and data is stored across the globe. In a traditional banking sense - you can think of these nodes as bank branches or stores that exist around the world.

This regions table contains the region_id and their respective region_name values

```{sql echo=FALSE, connection=con}
SELECT 
  *
FROM 
  dbo.regions
```

#### Table 2: Customer Nodes
Customers are randomly distributed across the nodes according to their region - this also specifies exactly which node contains both their cash and data.

This random distribution changes frequently to reduce the risk of hackers getting into Data Bank’s system and stealing customer’s money and data!

Below is a sample of the top 10 rows of the data_bank.customer_nodes

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM
  dbo.customer_nodes
```

#### Table 3: Customer Transactions
This table stores all customer deposits, withdrawals and purchases made using their Data Bank debit card.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM
  dbo.customer_transactions
```


### Case Study Questions
The following case study questions include some general data exploration analysis for the nodes and transactions before diving right into the core business questions and finishes with a challenging final request!

#### A. Customer Nodes Exploration

**1. How many unique nodes are there on the Data Bank system?**
```{sql, connection = con}
SELECT 
  COUNT(DISTINCT node_id) AS Unique_Nodes
FROM 
  dbo.customer_nodes
```
***

**2. What is the number of nodes per region?**
```{sql, connection = con}
SELECT 
  r.region_id, 
  r.region_name,
  COUNT(node_id) AS Number_of_nodes
FROM 
  dbo.customer_nodes n
JOIN 
  dbo.regions r
	  ON n.region_id = r.region_id
GROUP BY r.region_id, region_name 
ORDER BY region_id;
```
***

**3. How many customers are allocated to each region?**
```{sql, connection = con}
SELECT 
	region_id, 
	COUNT(DISTINCT customer_id) AS Number_of_Customers
FROM dbo.customer_nodes
GROUP BY region_id
ORDER BY region_id;
```
***

**4. How many days on average are customers reallocated to a different node?**
Not entirely sure that I understood the question. I interpret it as "*After* how many days...", rather than "How many *times*...", on average are they reallocated.
```{sql, connection = con}
WITH cte_day_diff AS (
	SELECT 
	customer_id, 
	node_id,
	start_date,
	end_date,
	DATEDIFF(day,start_date, end_date) AS diff
FROM	
	dbo.customer_nodes
WHERE end_date != '9999-12-31'
GROUP BY customer_id, node_id, start_date, end_date), 
cte_sum_diff AS (
	SELECT 
		customer_id, node_id, sum(diff) AS sum_diff
		FROM cte_day_diff
		GROUP BY customer_id, node_id)

SELECT 
  AVG(sum_diff) AS avg_reallocation_days
FROM cte_sum_diff; 
```
***

**5. What is the median, 80th and 95th percentile for this same reallocation days metric for each region?**
Re-used the same CTE from quesiton 4 and added another select query utilizing 
percentile_disc functions, returning a current value rather than an interpolated value. 
```{sql, connection = con}
WITH cte_day_diff AS (
	SELECT 
	region_id, 
	node_id,
	start_date,
	end_date,
	DATEDIFF(day,start_date, end_date) AS diff
FROM	
	dbo.customer_nodes
WHERE end_date != '9999-12-31'
GROUP BY region_id, node_id, start_date, end_date 
)

SELECT
	DISTINCT d.region_id,
	PERCENTILE_DISC(0.5) WITHIN GROUP (ORDER BY diff) 
	OVER (PARTITION BY d.region_id) AS Median,
	PERCENTILE_DISC(0.8) WITHIN GROUP (ORDER BY diff) 
	OVER (PARTITION BY d.region_id) AS eighty_centile,
	PERCENTILE_DISC(0.95) WITHIN GROUP (ORDER BY diff) 
	OVER (PARTITION BY d.region_id) AS ninetyfive_centile
FROM
	cte_day_diff AS d
	JOIN dbo.regions r
	ON 
		d.region_id = r.region_id
GROUP BY d.region_id, diff
ORDER BY d.region_id
```
***

#### B. Customer Transactions

**1. What is the unique count and total amount for each transaction type?**
```{sql, connection = con}
SELECT 
  txn_type, 
  COUNT(txn_type) AS count, 
  SUM(txn_amount) AS total_amount
FROM dbo.customer_transactions
GROUP BY txn_type
```
***

**2. What is the average total historical deposit counts and amounts for all customers?**
Making do with just a subquery over another CTE. 
```{sql, connection = con}
SELECT 
  AVG(t.counts) AS average_deposit_count, 
  AVG(t.sums) AS average_amount
FROM 
  (SELECT COUNT(txn_type) AS counts, AVG(txn_amount) AS sums
		FROM 
		  dbo.customer_transactions
		WHERE txn_type = 'deposit'
		GROUP BY customer_id) AS t
```
***

**3. For each month - how many Data Bank customers make more than 1 deposit and either 1 purchase or 1 withdrawal in a single month?**
Using case expression to create a counter for each transaction type and then 
issuing proper filter conditions.
```{sql, connection = con}
WITH month_transactions AS (
	SELECT 
	  MONTH(txn_date) AS month,
		customer_id, 
		SUM(CASE WHEN txn_type = 'deposit' THEN 0 ELSE 1 END) AS count_deposit,
		SUM(CASE WHEN txn_type = 'purchase' THEN 0 ELSE 1 END) AS count_purchase,
		SUM(CASE WHEN txn_type = 'withdrawal' THEN 0 ELSE 1 END) AS count_withdrawal
	FROM 
	  dbo.customer_transactions 
	GROUP BY customer_id, MONTH(txn_date))

SELECT 
  month, 
  COUNT(customer_id) AS count
FROM 
  month_transactions
WHERE count_deposit > 1
AND (count_purchase = 1 OR count_withdrawal = 1)
GROUP BY month;
```
***

**4. What is the closing balance for each customer at the end of the month?**
Could have utilized more CTE's for more readability, but I prefer it being more compact.
Note that I will revisit at a later point to fill in for months with no transactions!
```{sql, connection = con}
WITH cte_closing_balance AS (
	SELECT 
		customer_id,
		EOMONTH(txn_date) AS end_of_month,
		SUM(txn_amount*CASE WHEN txn_type = 'deposit' THEN 1
			ELSE -1 END) AS closing_balance,
		ROW_NUMBER() OVER (PARTITION BY customer_id, MONTH(txn_date) ORDER BY DAY(txn_date) DESC) AS txn_order
	FROM
		dbo.customer_transactions
	GROUP BY customer_id, txn_date)

SELECT 
	customer_id,
	end_of_month,
	closing_balance 
FROM 
  cte_closing_balance
WHERE txn_order = 1
ORDER BY customer_id, end_of_month
```
***

**5. What is the percentage of customers who increase their closing balance by more than 5%?**
Again, trying to minimize the amount of CTE's but I still ended up making 4 of them. 
We use DISTINCT in the end to get the percentage of customers who at least once 
experienced more than 5% growth.
```{sql, connection = con}
WITH cte_closing_balance AS (
	SELECT 
		customer_id,
		EOMONTH(txn_date) AS end_of_month,
		SUM(txn_amount*CASE WHEN txn_type = 'deposit' THEN 1
			ELSE -1 END) AS closing_balance,
		ROW_NUMBER() OVER (PARTITION BY customer_id, MONTH(txn_date) ORDER BY DAY(txn_date) DESC) AS txn_order
	FROM
		dbo.customer_transactions
	GROUP BY customer_id, txn_date),

cte_lead_balance AS(
SELECT
	customer_id,
	end_of_month,
	closing_balance,
	txn_order,
	LEAD(closing_balance) OVER (PARTITION BY customer_id ORDER BY end_of_month) AS next_balance
FROM 
  cte_closing_balance
WHERE txn_order = 1
),

cte_percent_growth AS (
SELECT
	customer_id,
	end_of_month,
	closing_balance,
	next_balance,
	FORMAT(CASE 
		WHEN closing_balance = 0 THEN NULL 
		WHEN closing_balance < 0 THEN ((CAST(next_balance AS FLOAT) - CAST(closing_balance AS FLOAT)) / CAST(closing_balance AS FLOAT)) * -1
		ELSE ((CAST(next_balance AS FLOAT)- CAST(closing_balance AS FLOAT)) / CAST(closing_balance AS FLOAT))
	END, 'P') AS percent_growth
FROM 
  cte_lead_balance
),

cte_filter AS (
SELECT
	customer_id,
	end_of_month,
	closing_balance,
	next_balance,
	percent_growth
FROM
	cte_percent_growth
GROUP BY customer_id, end_of_month, closing_balance, next_balance, percent_growth
HAVING percent_growth > FORMAT(5, 'P') -- We make another CTE as the prior one can't filter by the different case clauses.
AND percent_growth NOT LIKE '-%')

SELECT 
	FORMAT(ROUND(CAST(COUNT(DISTINCT customer_id) AS FLOAT) / (SELECT CAST(COUNT(DISTINCT customer_id) AS FLOAT)
FROM 
  cte_percent_growth),2), 'P') AS percentage_of_customers
FROM 
  cte_filter; 
```
***















































