---
title: "Data With Danny: Danny's Dinner"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```


```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "dwd#1", 
                      Trusted_Connection = "True")
```

### Introduction
Danny seriously loves Japanese food so in the beginning of 2021, he decides to embark upon a risky venture and opens up a cute little restaurant that sells his 3 favourite foods: sushi, curry and ramen.

Danny’s Diner is in need of your assistance to help the restaurant stay afloat - the restaurant has captured some very basic data from their few months of operation but have no idea how to use their data to help them run the business.

### Problem Statement
Danny wants to use the data to answer a few simple questions about his customers, especially about their visiting patterns, how much money they’ve spent and also which menu items are their favourite. Having this deeper connection with his customers will help him deliver a better and more personalised experience for his loyal customers.

He plans on using these insights to help him decide whether he should expand the existing customer loyalty program - additionally he needs help to generate some basic datasets so his team can easily inspect the data without needing to use SQL.

Danny has provided you with a sample of his overall customer data due to privacy issues - but he hopes that these examples are enough for you to write fully functioning SQL queries to help him answer his questions!

Danny has shared with you 3 key datasets for this case study:

* sales
* menu
* members

### Entity Relationship Diagram


### Datasets
All datasets exist within the dannys_diner database schema - be sure to include this reference within your SQL scripts as you start exploring the data and answering the case study questions.

#### Table 1: sales
The sales table captures all customer_id level purchases with an corresponding order_date and product_id information for when and what menu items were ordered.

```{sql echo=FALSE, connection=con, max.print=15}
SELECT
  *
FROM
  sales
```


#### Table 2: menu
The menu table maps the product_id to the actual product_name and price of each menu item.
```{sql echo=FALSE, connection=con}
SELECT
  *
FROM
  menu
```


#### Table 3: members
The final members table captures the join_date when a customer_id joined the beta version of the Danny’s Diner loyalty program.
```{sql echo=FALSE, connection=con}
SELECT
  *
FROM
  members
```


### Case Study Questions
Each of the following case study questions can be answered using a single SQL statement:

**1. What is the total amount each customer spent at the restaurant?**
```{sql, connection=con}
SELECT
	DISTINCT customer_id,
	SUM(price) AS total_amount
FROM
	sales
JOIN
	menu 
ON
	sales.product_id = menu.product_id
GROUP BY customer_id
```


***

**2. How many days has each customer visited the restaurant?**
```{sql, connection=con}
SELECT
	DISTINCT customer_id,
	COUNT(DISTINCT order_date) AS days
FROM
	sales
GROUP BY customer_id;
```
***

**3. What was the first item from the menu purchased by each customer?**
```{sql, connection=con}
WITH CTE_dr AS(
	SELECT 
		customer_id,
		order_date,
		product_name,
		DENSE_RANK() OVER(PARTITION BY customer_id
			ORDER BY order_date) AS dr
	FROM
		sales s
	JOIN
		menu m
	ON
		s.product_id = m.product_id
)

SELECT 
	customer_id,
	product_name
FROM
	CTE_dr
WHERE dr = 1
GROUP BY customer_id, product_name;
```
***

**4. What is the most purchased item on the menu and how many times was it purchased by all customers?**
```{sql, connection=con}
SELECT TOP 1
	product_name,
	(COUNT(s.product_id)) AS times_purchased
FROM
	sales s
JOIN
	menu m
ON
	s.product_id = m.product_id
GROUP BY product_name
ORDER BY times_purchased DESC
```
***

**5. Which item was the most popular for each customer?**
```{sql, connection=con}
WITH CTE_dr AS(
	SELECT 
		customer_id,
		product_name,
		COUNT(s.product_id) AS item_count,
		DENSE_RANK() OVER(PARTITION BY s.customer_id
			ORDER BY COUNT(s.customer_id) DESC) AS dr
	FROM
		sales s
	JOIN
		menu m
	ON
		s.product_id = m.product_id
	GROUP BY customer_id, product_name
)

SELECT 
	customer_id,
	product_name,
	item_count
FROM
	CTE_dr
WHERE dr = 1
GROUP BY customer_id, product_name, item_count;
```
***

**6. Which item was purchased first by the customer after they became a member?**
```{sql, connection=con}
WITH CTE_dr AS(
	SELECT 
		s.customer_id,
		join_date,
		order_date,
		product_id,
		DENSE_RANK() OVER(PARTITION BY s.customer_id
			ORDER BY order_date) AS dr
	FROM
		sales s
	JOIN
		members m
	ON
		s.customer_id = m.customer_id
	WHERE order_date >= join_date 
)

SELECT 
	customer_id,
	order_date,
	product_name
FROM
	CTE_dr dr
JOIN
	menu m
ON
	dr.product_id = m.product_id
WHERE dr = 1;
```
***

**7. Which item was purchased just before the customer became a member?**
```{sql, connection=con}
WITH CTE_dr AS(
	SELECT 
		s.customer_id,
		join_date,
		order_date,
		product_id,
		DENSE_RANK() OVER(PARTITION BY s.customer_id
			ORDER BY order_date DESC) AS dr
	FROM
		sales s
	JOIN
		members m
	ON
		s.customer_id = m.customer_id
	WHERE order_date < join_date 
)

SELECT 
	customer_id,
	order_date,
	product_name
FROM
	CTE_dr dr
JOIN
	menu m
ON
	dr.product_id = m.product_id
WHERE dr = 1;
```
***

**8. What is the total items and amount spent for each member before they became a member?**
```{sql, connection=con}
SELECT 
	DISTINCT s.customer_id,
	COUNT(DISTINCT s.product_id) AS total_items,
	SUM(price) AS total_expenditure
FROM
	sales s
JOIN
	members m
ON
	s.customer_id = m.customer_id
JOIN
	menu 
ON
	s.product_id = menu.product_id
WHERE order_date < join_date 
GROUP BY s.customer_id;
```
***

**9. If each $1 spent equates to 10 points and sushi has a 2x points multiplier - how many points would each customer have?**
```{sql, connection=con}
WITH CTE_points AS(
	SELECT
		customer_id,
		CASE 
			WHEN s.product_id = 1 THEN price*20
			ELSE price * 10
		END AS points
	FROM
		menu m
	JOIN
		sales s
	ON
		m.product_id = s.product_id
)
SELECT
	DISTINCT customer_id,
	SUM(points) AS total_points
FROM
	CTE_points
GROUP BY customer_id;

```
***

**10. In the first week after a customer joins the program (including their join date) they earn 2x points on all items, not just sushi - how many points do customer A and B have at the end of January?**
```{sql, connection=con}
WITH CTE_points AS(
	SELECT
		s.customer_id,
		order_date,
		s.product_id,
		CASE 
			WHEN s.product_id = 1 
			  OR (order_date BETWEEN join_date AND DATEADD(DAY, 7, join_date)) 
			  THEN price*20
			ELSE price * 10
		END AS points
	FROM
		menu m
	JOIN
		sales s
	ON
		m.product_id = s.product_id
	JOIN
		members mem
	ON
		s.customer_id = mem.customer_id
)
SELECT
	DISTINCT customer_id,
	SUM(points) AS total_points
FROM
	CTE_points
GROUP BY customer_id
```
Only difference to question 9 is that I added another OR condition with a between operator, which includes both the start and end value. I also added a parenthesis for better readability. 








