---
title: "Data With Danny: Balanced Tree Clothing Co."
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```

```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "dwd#7", 
                      Trusted_Connection = "True")
```

### Introduction
Balanced Tree Clothing Company prides themselves on providing an optimised range of clothing and lifestyle wear for the modern adventurer!

Danny, the CEO of this trendy fashion company has asked you to assist the team’s merchandising teams analyse their sales performance and generate a basic financial report to share with the wider business.

### Available Data
For this case study there is a total of 4 datasets for this case study - however you will only need to utilise 2 main tables to solve all of the regular questions, and the additional 2 tables are used only for the bonus challenge question!

#### Product Details
product_details includes all information about the entire range that Balanced Clothing sells in their store.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  product_details
```

#### Product Sales
sales contains product level information for all the transactions made for Balanced Tree including quantity, price, percentage discount, member status, a transaction ID and also the transaction timestamp.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM
  sales
```

### Case Study Questions
The following questions can be considered key business questions and metrics that the Balanced Tree team requires for their monthly reports.

Each question can be answered using a single query - but as you are writing the SQL to solve each individual problem, keep in mind how you would generate all of these metrics in a single SQL script which the Balanced Tree team can run each month.
 
#### A. High Level Sales Analysis

**1. What was the total quantity sold for all products?**
```{sql, connection = con}
SELECT 
	SUM(qty) AS total_quantity
FROM
	sales
```
***

**2. What is the total generated revenue for all products before discounts?**
```{sql, connection = con}
SELECT
	SUM(price*qty) AS Total_revenue
FROM 
	sales
```
***


**3. What was the total discount amount for all products?**
```{sql, connection = con}
SELECT
	SUM(qty*(price*(discount*0.01))) AS total_discount_amount
FROM
	sales
```
***

#### B. Transaction Analysis

**1. How many unique transactions were there?**
```{sql, connection = con}
SELECT
	COUNT(DISTINCT txn_id) AS count_unique_txn
FROM
	sales;
```
***

**2. What is the average unique products purchased in each transaction?**
```{sql, connection = con}
WITH cte_count AS(
	SELECT 
		COUNT(DISTINCT prod_id) AS unique_count
	FROM
		sales
	GROUP BY txn_id
)

SELECT
	AVG(unique_count) AS avg_unique_count
FROM 
	cte_count;
```
***

**3. What are the 25th, 50th and 75th percentile values for the revenue per transaction?**
```{sql, connection = con}
SELECT
	DISTINCT txn_id,
	PERCENTILE_DISC(0.25) WITHIN GROUP (ORDER BY qty*price-(qty*price*discount*0.01))
	OVER (PARTITION BY txn_id) AS twentyfive_centile,
	PERCENTILE_DISC(0.50) WITHIN GROUP (ORDER BY qty*price-(qty*price*discount*0.01))
	OVER (PARTITION BY txn_id) AS median,
	PERCENTILE_DISC(0.75) WITHIN GROUP (ORDER BY qty*price-(qty*price*discount*0.01))
	OVER (PARTITION BY txn_id) AS seventyfive_centile
FROM
    sales
GROUP BY txn_id, price, qty, discount
ORDER BY txn_id
```
***

**4. What is the average discount value per transaction?**
```{sql, connection = con}
SELECT
	txn_id,
	CAST(AVG(discount*price*qty*0.01) AS DECIMAL(10,2)) AS average_discount_value
FROM
	sales
GROUP BY txn_id
```
***

**5. What is the percentage split of all transactions for members vs non-members?**
```{sql, connection = con}
SELECT
	DISTINCT member,
	FORMAT(COUNT(1) * 100 / SUM(COUNT(1)) OVER() * 0.01, 'p') AS pct
FROM
	sales
GROUP BY member
```
***

**6. What is the average revenue for member transactions and non-member transactions?**
```{sql, connection = con}
SELECT
	DISTINCT member, 
	CAST(AVG(price*qty - (price*qty*discount*0.01)) AS DECIMAL(10,2)) AS avg_revenue
FROM
	sales
GROUP BY member;
```
***

#### C. Product Analysis

**1. What are the top 3 products by total revenue before discount?**
```{sql, connection = con}
WITH cte_total_revenue AS(
	SELECT
		prod_id,
		SUM(price*qty) AS total_revenue
	FROM
		sales
	GROUP BY prod_id
)

SELECT
	TOP 3 prod_id,
	total_revenue
FROM
	cte_total_revenue
ORDER BY total_revenue DESC;
```
***

**2. What is the total quantity, revenue and discount for each segment?**
```{sql, connection = con}
SELECT
	segment_name,
	SUM(qty) AS total_quantity,
	SUM(s.price*qty-(s.price*qty*discount*0.01)) AS total_revenue,
	SUM(s.price*qty*discount*0.01) AS total_discount
FROM sales s
	JOIN 
	product_details pd
	ON
		s.prod_id = pd.product_id
GROUP BY segment_name;
```
***

**3. What is the top selling product for each segment?**
```{sql, connection = con}
WITH cte_total_quantity AS(
	SELECT 
		segment_name,
		product_name,
		SUM(qty) AS total_quantity
	FROM
		sales s
	JOIN 
		product_details pd
	ON
		s.prod_id = pd.product_id
	GROUP BY segment_name, product_name
),

cte_rn AS(
	SELECT
		segment_name,
		product_name,
		total_quantity,
		ROW_NUMBER() OVER (PARTITION BY segment_name ORDER BY total_quantity DESC) AS rn
	FROM 
		cte_total_quantity
)

SELECT
	segment_name,
	product_name,
	total_quantity
FROM
	cte_rn
WHERE rn=1;
```
***

**4. What is the total quantity, revenue and discount for each category?**
```{sql, connection = con}
SELECT
	category_name,
	SUM(qty) AS total_quantity,
	SUM(qty*s.price-(qty*s.price*discount*0.01)) AS total_revenue,
	SUM(qty*s.price*discount*0.01) AS total_discount
FROM sales s
	JOIN 
	product_details pd
	ON
		s.prod_id = pd.product_id
GROUP BY category_name;
```
***

**5. What is the top selling product for each category?**
```{sql, connection = con}
WITH cte_total_quantity AS(
	SELECT 
		category_name,
		product_name,
		SUM(qty) AS total_quantity
	FROM
		sales s
	JOIN 
		product_details pd
	ON
		s.prod_id = pd.product_id
	GROUP BY category_name, product_name
),

cte_rn AS(
	SELECT
		category_name,
		product_name,
		total_quantity,
		ROW_NUMBER() OVER (PARTITION BY category_name ORDER BY total_quantity DESC) AS rn
	FROM 
		cte_total_quantity
)

SELECT
	category_name,
	product_name,
	total_quantity
FROM
	cte_rn
WHERE rn=1;
```
***

**6. What is the percentage split of revenue by product for each segment?**
```{sql, connection = con}
WITH cte_total_revenue AS(
	SELECT
		product_id,
		segment_name,
		product_name,
		SUM(SUM(s.price*qty - (s.price*qty*discount*0.01))) OVER (PARTITION BY segment_name) AS total_revenue
	FROM
		sales s
	JOIN 
		product_details pd
	ON
		s.prod_id = pd.product_id
	GROUP BY product_id, segment_name, product_name
)

SELECT
	segment_name,
	product_name,
	FORMAT(SUM(s.price*qty - (s.price*qty*discount*0.01)) / total_revenue,'P') AS split_pct_revenue
FROM
	cte_total_revenue tr
JOIN
	sales s
ON
	s.prod_id = tr.product_id
GROUP BY segment_name, product_name, total_revenue
ORDER BY segment_name;
```
Might later add a column with cumulative percentage making it easier to confirm percentages. 

***

**7. What is the percentage split of revenue by segment for each category?**
```{sql, connection = con}
WITH cte_total_revenue AS(
	SELECT
		product_id,
		category_name,
		segment_name,
		SUM(SUM(s.price*qty - (s.price*qty*discount*0.01))) OVER (PARTITION BY category_name) AS total_revenue
	FROM
		sales s
	JOIN 
		product_details pd
	ON
		s.prod_id = pd.product_id
	GROUP BY product_id, category_name, segment_name
)

SELECT
	category_name,
	segment_name,
	FORMAT(SUM(s.price*qty - (s.price*qty*discount*0.01)) / total_revenue,'P') AS split_pct_revenue
FROM
	cte_total_revenue tr
JOIN
	sales s
ON
	s.prod_id = tr.product_id
GROUP BY category_name, segment_name, total_revenue
ORDER BY category_name;
```
***

**8.  What is the percentage split of total revenue by category?**
```{sql, connection = con}
SELECT
	DISTINCT category_name,
	FORMAT(SUM(s.price*qty - (s.price*qty*discount*0.01)) OVER(PARTITION BY category_name) / 
	SUM(s.price*qty - (s.price*qty*discount*0.01)) OVER(), 'P') AS split_pct_revenue
FROM
	sales s
JOIN
	product_details pd
ON
	s.prod_id = pd.product_id;
```
***

**9. What is the total transaction “penetration” for each product? (hint: penetration = number of transactions where at least 1 quantity of a product was purchased divided by total number of transactions)**

To be continued