---
title: "Data With Danny: Clique Bait"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```

```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "dwd#6", 
                      Trusted_Connection = "True")
```

### Introduction
Clique Bait is not like your regular online seafood store - the founder and CEO Danny, was also a part of a digital data analytics team and wanted to expand his knowledge into the seafood industry!

In this case study - you are required to support Danny’s vision and analyse his dataset and come up with creative solutions to calculate funnel fallout rates for the Clique Bait online store.

### Available Data
For this case study there is a total of 5 datasets which you will need to combine to solve all of the questions.

**Users**

Customers who visit the Clique Bait website are tagged via their cookie_id.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  clique_bait.users
```

**Events**

Customer visits are logged in this events table at a cookie_id level and the event_type and page_id values can be used to join onto relevant satellite tables to obtain further information about each event.

The sequence_number is used to order the events within each visit.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  clique_bait.events
```


**Event Identifier**

The event_identifier table shows the types of events which are captured by Clique Bait’s digital data systems.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  clique_bait.event_identifier
```

**Campaign Identifier**

This table shows information for the 3 campaigns that Clique Bait has ran on their website so far in 2020.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  clique_bait.campaign_identifier
```


**Page Hierarchy**

This table lists all of the pages on the Clique Bait website which are tagged and have data passing through from user interaction events.

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  clique_bait.page_hierarchy
```

### Case Study Questions
**1. Digital Analysis**

Using the available datasets - answer the following questions using a single query for each one:

**1.How many users are there?**
```{sql connection=con}
SELECT 
	COUNT(DISTINCT user_id) AS nbr_users
FROM
	clique_bait.users;
```
***

**2.How many cookies does each user have on average?**
```{sql connection=con}
SELECT 
	AVG(count) AS avg_nbr_cookies
FROM
	(SELECT 
			user_id,
			COUNT(cookie_id) AS count
	FROM 
		clique_bait.users
	GROUP BY user_id) AS countd_cookies;
```
***

**3. What is the unique number of visits by all users per month?**
```{sql connection=con}
SELECT
	MONTH(event_time) AS month,
	COUNT(DISTINCT visit_id) AS unique_visits
FROM
	clique_bait.events
GROUP BY MONTH(event_time)
ORDER BY MONTH(event_time);
```
***

**4. What is the number of events for each event type?**
```{sql connection=con}
SELECT
	event_name,
	COUNT(*) AS nbr_events
FROM
	clique_bait.events ev 
JOIN
	clique_bait.event_identifier id
ON	ev.event_type = id.event_type
GROUP BY event_name
ORDER BY nbr_events DESC
```
***

**5. What is the percentage of visits which have a purchase event?**
```{sql connection=con}
SELECT
	FORMAT(0.01 * COUNT(DISTINCT visit_id)
	/ (SELECT COUNT(DISTINCT visit_id) FROM clique_bait.events), 'P') AS pct_purchase
FROM
	clique_bait.events
WHERE event_type=3
```
***

**6. What is the percentage of visits which view the checkout page but do not have a purchase event?**
```{sql connection=con}
WITH cte_checkout_purchase AS (
	SELECT
		MAX(CASE WHEN event_type = 3 THEN 1 ELSE 0 END) AS purchase,
		MAX(CASE WHEN page_id = 12 THEN 1 ELSE 0 END) AS checkout
	FROM 
		clique_bait.events
	GROUP BY visit_id
)

SELECT 
  FORMAT(CAST(SUM(checkout) - SUM(purchase) AS FLOAT) / SUM(checkout), 'P') as pct_visits_no_purchase
FROM cte_checkout_purchase;
```
***

**7. What are the top 3 pages by number of views?**
```{sql connection=con}
SELECT TOP 3 
	page_name,
	COUNT(*) as nbr_views
FROM
	clique_bait.events ev 
JOIN
	clique_bait.page_hierarchy ph
ON	ev.page_id = ph.page_id
GROUP BY page_name
ORDER BY nbr_views DESC;
```
***

**8. What is the number of views and cart adds for each product category?**
```{sql connection=con}
SELECT
	product_category,
	SUM(CASE WHEN event_type = 1 THEN 1 ELSE 0 END) AS views,
	SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS cart_adds
FROM
	clique_bait.events ev
JOIN
	clique_bait.page_hierarchy ph
ON
	ev.page_id = ph.page_id
WHERE product_category IS NOT NULL
GROUP BY product_category;
```
***

**9. What are the top 3 products by purchases?**
```{sql connection=con}
SELECT TOP 3
	page_name,
	SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS bought
FROM
	clique_bait.events ev1
JOIN
	clique_bait.page_hierarchy ph
ON
	ev1.page_id = ph.page_id
WHERE EXISTS (SELECT 1 FROM clique_bait.events ev2 WHERE ev1.visit_id = ev2.visit_id AND ev2.event_type=3)
GROUP BY page_name
ORDER BY bought DESC;
```
***

**2. Product Funnel Analysis**

Using a single SQL query - create a new output table which has the following details:

* **How many times was each product viewed?**
* **How many times was each product added to cart?**
* **How many times was each product added to a cart but not purchased (abandoned)?**
* **How many times was each product purchased?**

```{sql connection=con, immediate = TRUE}
WITH cte_t1 AS (
	SELECT
		page_name,
		SUM(CASE WHEN ev.event_type = 1 THEN 1 ELSE 0 END) AS product_view,
		SUM(CASE WHEN ev.event_type = 2 THEN 1 ELSE 0 END) AS added_to_cart
	FROM
		clique_bait.events ev
	JOIN
		clique_bait.page_hierarchy ph
	ON ev.page_id = ph.page_id
	WHERE product_id IS NOT NULL
	GROUP BY page_name
),

cte_bought AS (
	SELECT
		page_name,
		SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS bought
	FROM
		clique_bait.events ev1
	JOIN
		clique_bait.page_hierarchy ph
	ON ev1.page_id = ph.page_id
	WHERE EXISTS (SELECT 1 FROM clique_bait.events ev2 WHERE ev1.visit_id = ev2.visit_id AND ev2.event_type=3) 
		AND product_id IS NOT NULL
	GROUP BY page_name
),

cte_not_bought AS (
	SELECT
		page_name,
		SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS abandoned
	FROM
		clique_bait.events ev1
	JOIN
		clique_bait.page_hierarchy ph
	ON ev1.page_id = ph.page_id
	WHERE NOT EXISTS (SELECT 1 FROM clique_bait.events ev2 WHERE ev1.visit_id = ev2.visit_id AND ev2.event_type = 3) 
		AND product_id IS NOT NULL
	GROUP BY page_name
)

SELECT
	t.page_name AS product_name,
	product_view,
	added_to_cart,
	abandoned,
	bought

FROM
	cte_t1 t
JOIN
	cte_bought bt
	ON t.page_name = bt.page_name
JOIN
	cte_not_bought nbt
	ON t.page_name = nbt.page_name;
```
***

**Additionally, create another table which further aggregates the data for the above points but this time for each product category instead of individual products.**

```{sql connection=con, eval=F, echo=T}
SELECT
	product_category,
	SUM(product_view) AS views,
	SUM(added_to_cart) AS added_to_carts,
	SUM(abandoned) AS abandoned,
	SUM(bought) AS bought
FROM
	#temp_products tp
LEFT JOIN 
	clique_bait.page_hierarchy ph
	ON	tp.product_name = ph.page_name
GROUP BY product_category;
```
```{sql echo=FALSE, connection=con}
WITH cte_t1 AS (
	SELECT
		page_name,
		SUM(CASE WHEN ev.event_type = 1 THEN 1 ELSE 0 END) AS product_view,
		SUM(CASE WHEN ev.event_type = 2 THEN 1 ELSE 0 END) AS added_to_cart
	FROM
		clique_bait.events ev
	JOIN
		clique_bait.page_hierarchy ph
	ON ev.page_id = ph.page_id
	WHERE product_id IS NOT NULL
	GROUP BY page_name
),

cte_bought AS (
	SELECT
		page_name,
		SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS bought
	FROM
		clique_bait.events ev1
	JOIN
		clique_bait.page_hierarchy ph
	ON ev1.page_id = ph.page_id
	WHERE EXISTS (SELECT 1 FROM clique_bait.events ev2 WHERE ev1.visit_id = ev2.visit_id AND ev2.event_type=3) 
		AND product_id IS NOT NULL
	GROUP BY page_name
),

cte_not_bought AS (
	SELECT
		page_name,
		SUM(CASE WHEN event_type = 2 THEN 1 ELSE 0 END) AS abandoned
	FROM
		clique_bait.events ev1
	JOIN
		clique_bait.page_hierarchy ph
	ON ev1.page_id = ph.page_id
	WHERE NOT EXISTS (SELECT 1 FROM clique_bait.events ev2 WHERE ev1.visit_id = ev2.visit_id AND ev2.event_type = 3) 
		AND product_id IS NOT NULL
	GROUP BY page_name
),

meh AS(
  SELECT
  	t.page_name AS product_name,
  	product_view,
  	added_to_cart,
  	abandoned,
  	bought
  
  FROM
  	cte_t1 t
  JOIN
  	cte_bought bt
  	ON t.page_name = bt.page_name
  JOIN
  	cte_not_bought nbt
  	ON t.page_name = nbt.page_name
)
	
SELECT
	product_category,
	SUM(product_view) AS views,
	SUM(added_to_cart) AS added_to_carts,
	SUM(abandoned) AS abandoned,
	SUM(bought) AS bought
FROM
	meh tp
LEFT JOIN 
	clique_bait.page_hierarchy ph
	ON	tp.product_name = ph.page_name
GROUP BY product_category;
```

I turned the previous table into a temporary table called ***#temp_products*** so I did not have to repeat all the code.

***

Use your 2 new output tables - answer the following questions:

**Which product had the most views, cart adds and purchases?**
```{sql connection=con, eval=F, echo=T}
SELECT TOP 1
	(SELECT TOP 1 product_name FROM #temp_products ORDER BY product_view DESC) AS most_views,
	(SELECT TOP 1 product_name FROM #temp_products ORDER BY added_to_cart DESC) AS most_added_to_cart,
	(SELECT TOP 1 product_name FROM #temp_products ORDER BY bought DESC) AS most_bought
FROM
	#temp_products;
```
***

**Which product was most likely to be abandoned?**
```{sql connection=con, eval=F, echo=T}
SELECT TOP 1
	product_name AS most_abandoned
FROM
	#temp_products
ORDER BY abandoned DESC;
```
***

**Which product had the highest view to purchase percentage?**
```{sql connection=con, eval=F, echo=T}
SELECT TOP 1 
	product_name,
	FORMAT(ROUND(CAST(bought AS FLOAT)/product_view,3), 'P') AS view_to_purchase_pct
FROM
	#temp_products
ORDER BY view_to_purchase_pct DESC;
```
***

**What is the average conversion rate from view to cart add?**
```{sql connection=con, eval=F, echo=T}
SELECT
	FORMAT(ROUND(AVG(CAST(added_to_cart AS FLOAT)/product_view),3), 'P') AS avg_cart_add_conversion
FROM
	#temp_products;
```
***

**What is the average conversion rate from cart add to purchase?**
```{sql connection=con, eval=F, echo=T}
SELECT
	FORMAT(ROUND(AVG(CAST(bought AS FLOAT)/added_to_cart),3), 'P') AS avg_purchase_conversion
FROM
	#temp_products;
```
***

**3. Campaigns Analysis**

Generate a table that has 1 single row for every unique visit_id record and has the following columns:

* **user_id**
* **visit_id**
* **visit_start_time**: the earliest event_time for each visit
* **page_views**: count of page views for each visit
* **cart_adds**: count of product cart add events for each visit
* **purchase**: 1/0 flag if a purchase event exists for each visit
* **campaign_name**: map the visit to a campaign if the visit_start_time falls between the start_date and end_date
* **impression**: count of ad impressions for each visit
* **click**: count of ad clicks for each visit
* **(Optional column) cart_products**: a comma separated text value with products added to the cart sorted by the order they were added to the cart (hint: use the sequence_number)

```{sql connection=con}
SELECT
	user_id,
	visit_id,
	MIN(event_time) AS visit_start_time,
	SUM(1) AS page_views,
	SUM(CASE WHEN event_type=2 THEN 1 ELSE 0 END)  AS cart_adds,
	MAX(CASE WHEN event_type=3 THEN 1 ELSE 0 END) AS purchase,
	c_id.campaign_name,
	SUM(CASE WHEN event_type=4 THEN 1 ELSE 0 END) AS impression,
	SUM(CASE WHEN event_type=5 THEN 1 ELSE 0 END) AS click,
	STRING_AGG(CASE 
	              WHEN ph.product_id IS NOT NULL AND ev.event_type=2 
	              THEN ph.page_name ELSE NULL END, ', ') 
	              WITHIN GROUP (ORDER BY sequence_number) AS cart_products
FROM
	clique_bait.events ev
JOIN
	clique_bait.users u
	ON 
		ev.cookie_id = u.cookie_id
JOIN
	clique_bait.campaign_identifier c_id
	ON
		ev.event_time BETWEEN c_id.start_date AND c_id.end_date
JOIN
	clique_bait.page_hierarchy ph
	ON
		ev.page_id = ph.page_id
GROUP BY user_id, visit_id, campaign_name
```
Pretty clean. I first planned to filter everything on first sequence_number by adding another join, and just SUM by partition of visit_id for each column, but STRING_AGG() wasn't "a valid windowing function" and it "cannot be used with the OVER() clause", so I settled on using group by instead and removed the partitions. I also learnt that it's possible to join tables by time !







