---
title: "Predicting Default for Refinance Loan Applications"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---

# Introduction
In this dataset you will find consumer loan applications and the outcome flag for these
applications (default flag).

**Task:** Build a logistic regression model that predicts probability of default
for **refinance loans.** Explain and present your model, explain the selection 
of variables, and discuss the data limitations. Explain what can be improved in
your model and discuss data limitations. 

## Data

Below is a description of the dataset variables.

* **ApplicationDate** - Date of application (MMYYYY)
* **Blanco_loan_debt** -  Blanco loan debt (SEK)
* **Bureau_Score** - Credit score from the bureau
* **CivilStatus** - Married/Single/Cohabiting
* **Has_Co_Appl_Flag** - 1 if co-applicant exist, else 0
* **Has_Mortgage** - 1 if applicant has a mortgage loan
* **NrOfCars** - Number of cars registered on the applicant
* **NrOfChildren** - Number of children
* **NumberOfLoans** - Number of loans (all loan types)
* **UC_Req_L12M** - Number of loan requests done past 12m
* **UC_Req_L3M** - Number of loan requests done past 6m
* **UC_Req_L6M** - Number of loan requests done past 3m
* **age** - Age
* **broker_id** - Id of the broker that the applicant used when applying
* **default_flag** -  1= default, 0= not defaulted within 12m from application date
* **gender** - "F"=Female, "M"=Male
* **has_co_adress** - 1 if applicant has a C/O-address, else 0
* **has_creditcard_loan** - 1 if applicant has at least one credit card loan, else 0
* **has_repayment_loan** - 1 if applicant has at least one repayment loan, else 0
* **income_year1** - Gross income most recent income year
* **income_year2** - Gross income past income year
* **requested_amount** - Total amount requested to loan
* **refinance_flag** - 1= Refinance loan, 0 = Not a refinance loan

In short, we have to *filter by refinance_flag = 1* to keep our analysis limited to
refinance loans, and predict the dependent binary variable **default_flag** in 
relation to the other variables.

## Data Wrangling
We'll load a few necessary packages beforehand, tidyverse and lubridate, 
knowing that we'll most certainly work with data manipulation and date types. 
```{r Loading packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

Once we have all the libraries necessary, we can read the data in from CSV file.
```{r Loading data, message=FALSE, warning=FALSE}
theme_set(theme_light())
df.loan_apps <- read.csv("case_data.csv", sep=";")
glimpse(df.loan_apps)
```

The data consists of 25000 rows with 23 columns. By observe that several columns 
need to be converted; the applicationDate to date format, the CivilStatus and 
gender to factors, each flag -and "has" variable to logical data types, 
and finally remove the dollar signs from previous years gross income and convert
to numeric.

```{r}
df.clean <- df.loan_apps %>% 
  mutate(ApplicationDate = format(parse_date_time(ApplicationDate, orders = c("m/Y")), "%m/%Y"),
         CivilStatus = as.factor(CivilStatus),
         gender = as.factor(gender),
         across(starts_with("has"), as.logical),
         across(ends_with("flag"), as.logical),
         across(starts_with("income_year"), 
                ~gsub("[$,]", "", .) %>% 
                as.numeric))
glimpse(df.clean)
```

Are there any obvious data augmentation to be made? Similar to credit scores, 
there are other ratios that I presume could lessen the number constraints on the
model, such as debt-to-income, loan-to-value, annual percentage rate, etc. However, 
I must confess that I am ill informed in this area and I would much rather work
with what we got. It might as well prove to be redundant. As such, we'll stick
with the current set of variables.

## Exploratory data analysis
Let us begin by having a look at the number of missing values:
```{r}
sapply(df.clean, function(x) sum(is.na(x)))
```
The variable income_year1 is the only one with missing values, reassuring us 
that values of 0 within the data are not to be confused with actual missing values.
The question now is whether or not we should remove rows with missing values or perhaps 
remove income_year1 entirely, since it could very likely be correlated with 
income_year2.

Let's remove the 1224 rows with missing values but also take a peek at 
the correlation matrix of our data.
```{r message=FALSE, warning=FALSE}
df.refinance <- df.clean %>% filter(refinance_flag==TRUE) %>%
  na.omit() %>% 
  dplyr::select(-refinance_flag)

df.refinance %>% select(-ApplicationDate) %>% 
  mutate(gender=as.numeric(gender),
         CivilStatus=as.numeric(CivilStatus)) %>% 
  cor() %>% ggcorrplot::ggcorrplot(lab=TRUE,
                                   lab_size = 3,
                                   digits = 1, 
                                   type = "lower",
                                   ggtheme = ggplot2::theme_void,
                                   tl.cex = 8)
```

We find that both the number of requests for loans across past months and the 
past reported incomes have considerable high correlations. I'm hesitant to 
remove either variables since we are not dealing with perfect collinearity,
but it might be an option for improving our future model. It's also worth noting 
that certain models can "infer" this in feature selection and remove useless variables.

Let's take a quick look at our dependent variable:
```{r}
df.refinance %>% group_by(default_flag) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(default_flag, count, fill=default_flag, color = default_flag)) +
  geom_bar(stat="identity", alpha=0.4)

prop.table(table(df.refinance$default_flag))
```

This is an issue. Having big imbalances in the data wont provide our model with enough
neccessary information about the minority class, in this case for the 
applications that end up defaulting, as the model strives to minimize 
error in which the minority class contributes very little, resulting in 
biased predictions and misleading accuracies.

As a fix, let's perform random undersampling! It's a method for reducing the number of 
observations from the majority class as a way to balance out the dataset.
```{r}
t <- which(df.refinance$default_flag==TRUE)
f <- which(df.refinance$default_flag==FALSE)
f.downsample <- sample(f, length(t))
df.refinance_down <- df.refinance[c(f.downsample, t),]

df.refinance_down %>% group_by(default_flag) %>% 
  summarise(count = n())
prop.table(table(df.refinance_down$default_flag))
```

There are also "Informative" undersampling methods with both an unsupervised and 
supervised learning algorithm, but we'll settle for the quick, random undersampling method.

## Analysis
My first thought was to perform a stepwise logistic regression model for 
obtaining the best variables, since that is what I'm most familiar with from my
time studying statistics. However, I've found multiple online resources suggesting 
that stepwise model selection is a cause for overfitting, by having coefficients
that are too large and resulting in a highly biased outcome with nonreliable accuracy. 

Now since we do have a few correlated variables, a down sampled dataset, and are asked to perform and also 
motivate feature selection, **using LASSO** would be a good choice. It is a Regularization method 
which penalizes large coefficients and can specifically perform estimation on very small sample sizes, 
implement cross validation for tuning a hyperparameter, as a way for 
balancing bias and variance, and computes a simpler model by forcing coefficients
of lesser contributive variables to be zero. 

We split the data into a training set (75%) for estimating parameters and another
set reserved for testing and evaluation (25%). 
```{r message=FALSE, warning=FALSE}
set.seed(2022)
library(tidymodels)
library(glmnet)
library(caret)

split <- initial_split(df.refinance_down)
train <- training(split)
test <- testing(split)

x.train <- model.matrix(default_flag~., train)[,-15]
y.train <- train$default_flag

x.test <- model.matrix(default_flag~., test)[,-15]
y.test <- test$default_flag
```

LASSO is a penalized regression which utilises a constant lambda for adjusting 
the amount of coefficient shrinkage, by optimizing lambda through iterations 
we can minimize the cross validation prediction error rate. We'll use the 
default value of 10 blocks for cross validation.
```{r}
cv.lasso <- cv.glmnet(x.train, y.train, family = "binomial", alpha = 1, nfolds = 10, keep = TRUE)
plot(cv.lasso)
```

The left dashed vertical line is the optimal log value of lambda for minimizing 
prediction error, and the right one is for the simplest model within one 
standard error of the optimal model, in other words the best balance between 
accuracy and simplicity. The values are:
```{r}
cv.lasso$lambda.min
cv.lasso$lambda.1se
```

We proceed to fit the final models with their respective lambda values onto the 
training data to find which regression coefficients are best.
```{r}
model.1se <- glmnet(x.train, y.train, alpha = 1, family = "binomial", 
                lambda = cv.lasso$lambda.1se)
coef(model.1se)
```
There are 16 coefficients remaining, the rest have been shrunken down to 0. Amongst 
these coefficients, increasing the likelihood of defaulting on a 
refinance loan is dependent on the requested amount, if you are single, if you have requested 
a loan in recent months, and if you are a male. The remaining coefficients decrease the 
likelihood on defaulting which pertain mostly to people who have grown old, 
started a family, and who are financially stable enough to having already 
acquired several loans under their name. There are also a few select months that affect 
the predictions in both ways.

```{r}
model.min <- glmnet(x.train, y.train, alpha = 1, family = "binomial",
                lambda = cv.lasso$lambda.min)
coef(model.min)
```

This is the model with the smallest achieved MSE which is also considered the most
accurate model. It's likely similar to a full model in terms of accuracy since 
LASSO only shrinks useless variables and it is close to being completely unrestricted.

Let's create predictions onto the test data, by using the fitted probabilities, and examine
model accuracy and the confusion matrix.
```{r}
# fitting predictions
prob.min <- predict(model.min, newx = x.test, type = "response")
prob.1se <- predict(model.1se, newx = x.test, type = "response")

pred.min <- ifelse(prob.min > 0.5, "TRUE", "FALSE")
pred.1se <- ifelse(prob.1se > 0.5, "TRUE", "FALSE")

# Model accuracy
mean(pred.min == test$default_flag)

mean(pred.1se == test$default_flag)
```

The models differ by 2%, having 68.6% and 66.6% accuracy respectively, 
but the .1se model is simpler with a lot fewer predictors. 


```{r}
# confusion matrix for the .se1 model
cnf <- cv.glmnet(x.train, y.train, alpha = 1, family = "binomial")
confusion.glmnet(cnf, newx = x.test, newy = y.test)
```

The confusion matrix tells us how many instances whose correct default status
was missclassified for the other. In binary classification, the diagonal 
entries are commonly referred to as the true positives and the true negatives, 
and the other two are the false positives and false negatives.

## Discussing models and data limitations

Generic takes; LASSO is better for feature selection or sparse model selection. Ridge 
regression may give better prediction since it uses all variables. If an outcome 
is better predicted by many weak predictors, then ridge regression or 
bagging/boosting will outperform both forward stepwise regression and LASSO by a
long shot. LASSO is much faster than forward stepwise regression.

Concerning whether or not income_year1 and income_year2 are in fact reported in 
dollars or just a typo, I would personally assume that it is in fact a typo 
since the other variables are reported in SEK which would be more in line with
the other figures in terms of value, however, I do confess that I lack the 
necessary domain knowledge and would normally consult someone else in a real 
life scenario. If they are in fact reported correctly I would consider it a data
limitation, as I assume having variables with different currencies makes it no 
longer comparable since they are on different scales, and I would instead implement
a column for conversion rates between the two currency at the time of date, which
I would also drill down in the applicationDate field.

Other shortcomings; the data is missing interest rate, as blanco loans 
are usually higher with a wide range that likely has a predictive 
power. Similarly, a loan term could also be useful as it determines not only 
how long the borrower will be in debt, but how high the borrower’s monthly
loan payments and overall loan costs will be. Another variable that might prove 
helpful, but generic, is education level of the applicant, but also of the co-applicant 
since we are already have a flag for it. Might as well add co-applicant income.











