---
title: "Data With Danny: Foodie-Fi"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```


```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "dwd#3", 
                      Trusted_Connection = "True")
```

### Introduction
Subscription based businesses are super popular and Danny realised that there was a large gap in the market - he wanted to create a new streaming service that only had food related content - something like Netflix but with only cooking shows!

Danny finds a few smart friends to launch his new startup Foodie-Fi in 2020 and started selling monthly and annual subscriptions, giving their customers unlimited on-demand access to exclusive food videos from around the world!

Danny created Foodie-Fi with a data driven mindset and wanted to ensure all future investment decisions and new features were decided using data. This case study focuses on using subscription style digital data to answer important business questions.

### Available Data
Danny has shared the data design for Foodie-Fi and also short descriptions on each of the database tables - our case study focuses on only 2 tables but there will be a challenge to create a new table for the Foodie-Fi team.

All datasets exist within the foodie_fi database schema - be sure to include this reference within your SQL scripts as you start exploring the data and answering the case study questions.

#### Entity Relationship Diagram


#### Table 1: plans
Customers can choose which plans to join Foodie-Fi when they first sign up.

Basic plan customers have limited access and can only stream their videos and is only available monthly at $9.90

Pro plan customers have no watch time limits and are able to download videos for offline viewing. Pro plans start at $19.90 a month or $199 for an annual subscription.

Customers can sign up to an initial 7 day free trial will automatically continue with the pro monthly subscription plan unless they cancel, downgrade to basic or upgrade to an annual pro plan at any point during the trial.

When customers cancel their Foodie-Fi service - they will have a churn plan record with a null price but their plan will continue until the end of the billing period.

```{sql echo=FALSE, connection=con}
SELECT *
FROM plans
```


#### Table 2: subscriptions
Customer subscriptions show the exact date where their specific plan_id starts.

If customers downgrade from a pro plan or cancel their subscription - the higher plan will remain in place until the period is over - the start_date in the subscriptions table will reflect the date that the actual plan changes.

When customers upgrade their account from a basic plan to a pro or annual pro plan - the higher plan will take effect straightaway.

When customers churn - they will keep their access until the end of their current billing period but the start_date will be technically the day they decided to cancel their service.

```{sql echo=FALSE, connection=con}
SELECT 
	customer_id,
	plan_name,
	price,
	start_date
FROM
	plans p
	JOIN
	subscriptions s
	ON
		p.plan_id = s.plan_id
WHERE customer_id IN (1,2,11,13,15,16,18,19)
```

### Case Study Questions
This case study is split into an initial data understanding question before diving straight into data analysis questions before finishing with 1 single extension challenge.

#### A. Customer Journey
Based off the 8 sample customers provided in the sample from the subscriptions table, write a brief description about each customer’s onboarding journey.

Try to keep it as short as possible - you may also want to run some sort of join to make your explanations a bit easier!

```{sql echo=FALSE, connection=con}
SELECT 
	customer_id,
	plan_name,
	price,
	start_date
FROM
	plans p
	JOIN
	subscriptions s
	ON
		p.plan_id = s.plan_id
WHERE customer_id = 1
```
Customer id 1 started trial on 1st of August 2020 and afterwards paid for a basic monthly subscription on August the 8th 2020.

```{sql echo=FALSE, connection=con}
SELECT 
	customer_id,
	plan_name,
	price,
	start_date
FROM
	plans p
	JOIN
	subscriptions s
	ON
		p.plan_id = s.plan_id
WHERE customer_id = 11
```
Customer id 11 started trial on 19th of november 2020 and let it run out with no renewal.

```{sql echo=FALSE, connection=con}
SELECT 
	customer_id,
	plan_name,
	price,
	start_date
FROM
	plans p
	JOIN
	subscriptions s
	ON
		p.plan_id = s.plan_id
WHERE customer_id = 15
```
Customer id 15 started trial on 17 of March 2020, upgraded to a pro monthly subscription right after which has added benefits, then canceled his subscription on April the 29th, having access to his account until 24th of May.

#### B. Data Analysis Questions

**1. How many customers has Foodie-Fi ever had?**
```{sql, connection = con}
SELECT
	COUNT(DISTINCT customer_id)
FROM
	subscriptions
```
***

**2. What is the monthly distribution of trial plan start_date values for our dataset - use the start of the month as the group by value**
```{sql, connection = con, max.print = 12}
SELECT
	MONTH(start_date) AS month_of_year,
	COUNT(plan_id) AS count
FROM 
	subscriptions
WHERE plan_id = 0
GROUP BY MONTH(start_date) 
ORDER BY MONTH(start_date)
```
***

**3. What plan start_date values occur after the year 2020 for our dataset? Show the breakdown by count of events for each plan_name**
```{sql, connection = con}
SELECT 
	plan_name,
	COUNT(plan_name) AS count_2021
FROM 
	subscriptions s
	RIGHT JOIN
	plans p
	ON
	s.plan_id = p.plan_id
WHERE start_date >= '2021-01-01'
GROUP BY plan_name
```
Trial have 0 occurences and does not show in the table.

***

**4. What is the customer count and percentage of customers who have churned rounded to 1 decimal place?**
```{sql, connection = con}
SELECT 
	COUNT(customer_id) AS count,
	ROUND(CAST(100 * COUNT(customer_id) AS FLOAT) / (
		SELECT 
			count(DISTINCT customer_id)
		FROM
			subscriptions),1) AS pct
FROM
	subscriptions
WHERE plan_id = 4
```
***


**5. How many customers have churned straight after their initial free trial - what percentage is this rounded to the nearest whole number?**
```{sql, connection = con}
WITH cte_seq AS (
	SELECT 
		plan_id,
		lead(plan_id) OVER (order by customer_id) AS next_plan
	FROM 
		subscriptions
)

SELECT
	COUNT(*) AS count,
	ROUND(CAST(100*COUNT(*) AS FLOAT) / (
		SELECT
			COUNT(DISTINCT customer_id)
		FROM 
			subscriptions),1) AS pct
FROM
	cte_seq
WHERE plan_id = 0
AND next_plan = 4

```
***

**6. What is the number and percentage of customer plans after their initial free trial?**
```{sql, connection = con}
WITH cte_seq AS (
	SELECT 
		plan_id,
		lead(plan_id) OVER (order by customer_id) AS next_plan
	FROM 
		subscriptions
)

SELECT
	next_plan AS new_plan,
	COUNT(*) AS count,
	ROUND(CAST(100*COUNT(*) AS FLOAT) / (
		SELECT
			COUNT(DISTINCT customer_id)
		FROM 
			subscriptions),1) AS pct
FROM
	cte_seq
WHERE plan_id = 0
GROUP BY next_plan;
```
***

**7. What is the customer count and percentage breakdown of all 5 plan_name values at 2020-12-31?**
```{sql, connection = con}
WITH cte_seq AS (
	SELECT 
		customer_id,
		plan_id,
		start_date,
		lead(start_date) OVER(PARTITION BY customer_id ORDER BY start_date) AS next_date
	FROM 
		subscriptions
	WHERE start_date <= '2020-12-31' -- To cut-off before 2021.
) 

SELECT
	plan_id,
	COUNT(DISTINCT customer_id) AS count,
	ROUND(CAST(100* COUNT(DISTINCT customer_id) AS FLOAT) /
		(SELECT count(DISTINCT customer_id) 
		 FROM subscriptions),1) AS pct
FROM
	cte_seq
WHERE next_date IS NULL -- To get the final subscription plan for each distinct customer.
GROUP BY plan_id;
```
***


**8. How many customers have upgraded to an annual plan in 2020?**
```{sql, connection = con}
SELECT
	COUNT(DISTINCT customer_id) AS count
FROM subscriptions
WHERE start_date <= '2020-12-31'
AND plan_id = 3
```
***

**9. How many days on average does it take for a customer to an annual plan from the day they join Foodie-Fi?**
```{sql, connection = con}
WITH cte_join_day AS(
	SELECT 
		DISTINCT customer_id,
		start_date AS join_date
	FROM
		subscriptions
	WHERE plan_id=0
),
cte_annual_day AS(
	SELECT
		DISTINCT customer_id,
		start_date AS annual_date
	FROM
		subscriptions
	WHERE plan_id=3
)

SELECT
	AVG(DATEDIFF(DAY, join_date, annual_date)) AS avg_days_to_upgrade
FROM cte_join_day t
JOIN cte_annual_day a
  ON t.customer_id = a.customer_id;
```
***

**10. Can you further breakdown this average value into 30 day periods (i.e. 0-30 days, 31-60 days etc)**
```{sql, connection = con}
WITH cte_join_day AS(
	SELECT 
		DISTINCT customer_id,
		start_date AS join_date
	FROM
		subscriptions
	WHERE plan_id=0
),
cte_annual_day AS(
	SELECT
		DISTINCT customer_id,
		start_date AS annual_date
	FROM
		subscriptions
	WHERE plan_id=3
),

cte_count_days AS(
	SELECT
		DATEDIFF(DAY, join_date, annual_date) AS days_to_upgrade
	FROM cte_join_day t
	JOIN cte_annual_day a
		ON t.customer_id = a.customer_id
)

SELECT
	COUNT(CASE WHEN days_to_upgrade <31 THEN 1 END) AS '0-30',
	COUNT(CASE WHEN days_to_upgrade > 30 AND days_to_upgrade < 61 THEN 1 END) AS '31-60',
	COUNT(CASE WHEN days_to_upgrade > 60 AND days_to_upgrade < 91 THEN 1 END) AS '61-90',
	COUNT(CASE WHEN days_to_upgrade > 90 AND days_to_upgrade < 121 THEN 1 END) AS '91-120',
	COUNT(CASE WHEN days_to_upgrade > 120 AND days_to_upgrade < 151 THEN 1 END) AS '121-150',
	COUNT(CASE WHEN days_to_upgrade > 150 AND days_to_upgrade < 181 THEN 1 END) AS '151-180',
	COUNT(CASE WHEN days_to_upgrade > 180 AND days_to_upgrade < 211 THEN 1 END) AS '181-210',
	COUNT(CASE WHEN days_to_upgrade > 210 AND days_to_upgrade < 241 THEN 1 END) AS '211-240',
	COUNT(CASE WHEN days_to_upgrade > 240 AND days_to_upgrade < 271 THEN 1 END) AS '241-270',
	COUNT(CASE WHEN days_to_upgrade > 270 AND days_to_upgrade < 301 THEN 1 END) AS '271-300',
	COUNT(CASE WHEN days_to_upgrade > 300 AND days_to_upgrade < 331 THEN 1 END) AS '301-330',
	COUNT(CASE WHEN days_to_upgrade > 330 THEN 1 END) AS '331-360'
FROM
	cte_count_days;
```
A simple but tedious solution to hardcode each bin. Could also do it in a long format with 3 columns but it requires an additional CTE:

```{sql, connection=con}
WITH cte_join_day AS(
	SELECT 
		DISTINCT customer_id,
		start_date AS join_date
	FROM
		subscriptions
	WHERE plan_id=0
),
cte_annual_day AS(
	SELECT
		DISTINCT customer_id,
		start_date AS annual_date
	FROM
		subscriptions
	WHERE plan_id=3
),

cte_count_days AS(
	SELECT
		DATEDIFF(DAY, join_date, annual_date) AS days_to_upgrade
	FROM cte_join_day t
	JOIN cte_annual_day a
		ON t.customer_id = a.customer_id
),

cte_day_bracket AS(
	SELECT 
		CASE 
			WHEN days_to_upgrade <31 THEN '0-30'
			WHEN days_to_upgrade > 30 AND days_to_upgrade < 61 THEN  '31-60'
			WHEN days_to_upgrade > 60 AND days_to_upgrade < 91 THEN  '61-90'
			WHEN days_to_upgrade > 90 AND days_to_upgrade < 121 THEN  '91-120'
			WHEN days_to_upgrade > 120 AND days_to_upgrade < 151 THEN  '121-150'
			WHEN days_to_upgrade > 150 AND days_to_upgrade < 181 THEN  '151-180'
			WHEN days_to_upgrade > 180 AND days_to_upgrade < 211 THEN  '181-210'
			WHEN days_to_upgrade > 210 AND days_to_upgrade < 241 THEN  '211-240'
			WHEN days_to_upgrade > 240 AND days_to_upgrade < 271 THEN  '241-270'
			WHEN days_to_upgrade > 270 AND days_to_upgrade < 301 THEN  '271-300'
			WHEN days_to_upgrade > 300 AND days_to_upgrade < 331 THEN  '301-330'
			WHEN days_to_upgrade > 330 THEN '331-360'
			ELSE 'NA'
		END AS day_bracket,
		days_to_upgrade
	FROM
		cte_count_days
)

SELECT
    day_bracket,
	COUNT(days_to_upgrade) AS count
FROM
    cte_day_bracket
GROUP BY
    day_bracket
ORDER BY day_bracket;
```
**Reminder** I need to order it

***

**11. How many customers downgraded from a pro monthly to a basic monthly plan in 2020?**
```{sql, connection = con}
WITH cte_seq AS (
	SELECT 
		plan_id,
		lead(plan_id) OVER (PARTITION BY customer_id ORDER BY customer_id) AS next_plan
	FROM 
		subscriptions
)

SELECT
	COUNT(*) AS count
FROM
	cte_seq
WHERE plan_id = 2
AND next_plan = 1
-- 0.

```
***








