---
title: "Cyclistic Case Study"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  github_document
---
```{r, include=FALSE}
knitr::opts_chunk$set(fig.align="center", figure.width=14) 
```
### Preamble
This report is to show my approach and work in solving a case study provided by 
Google Data Analytics Course. While there are uses for SQL and Tableau in
this report, I'll leave those for future case studies and instead only
focus on R and Rmarkdown. This will be a good opportunity to get accustomed
with the tidyverse package while also describing my thought process with
proper reasoning, comments and visualizations. 

### Background 

Cyclistic is a bike-share program that features more than 5,800 bicycles and 600 docking stations. Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.

There are three different pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members. **Aid the marketing team in converting casual bike riders into annual members, 
by profiling each group and giving recommendations.**

### Preparations
This report uses a fictional, publicly available dataset on bike trips in Chicago. 
The data was accessed on March the 1st 2022, by the following link
https://divvy-tripdata.s3.amazonaws.com/index.html.

Loading packages suitable for manipulation of data, time and coordinates, 
as well as useful packages for observing the data.
```{r Loading packages, message=FALSE, warning=FALSE, include=TRUE}
library(tidyverse) # tidy data & piping
library(lubridate) # 
library(geosphere)
library(scales)
library(ggpubr)
```

Starting off, the files are found too big and bothersome to load through excel. 
```{r File info}
files <- list.files("biketrip_datafiles", full.names = TRUE)
paste(sum(file.size(files)),"bytes")
```
The files together are above 1GB, making it slow and tedious to clean and manipulate 
through excel or google sheets. It's a big reason for utilizing R instead of spreadsheets.
The data is contained within monthly data frames having same number of columns 
and same respective data type. As such, concatenating the csv files can be done
through the following code.
```{r Import data, message=FALSE, warning=FALSE, include=TRUE}
# applies read_csv to every element and returns a list, binding together the rows into 1.
df_trips <- files %>% 
  lapply(read_csv) %>% 
  bind_rows 
```

Getting an overview of the data.
```{r Data glimpse}
df_trips %>% glimpse()
```
There are over 5.6 million observations for 13 columns. Each variable seems to be 
stored in reasonable data types nor are there any unnecessary white spaces or 
irregular names in need of trimming. Manipulating and wrangling the data, with
the stakeholders in mind, enables it to be used and analysed. 
Therefore, checking for missing values and duplicates is in order. 

```{r NAs and duplicates}
sum(duplicated(df_trips$ride_id))
sapply(df_trips, function(y) sum(is.na(y))) # https://stackoverflow.com/a/31145812
```
There are no duplicates present in the data frame, but station names and ID's do have
equal number of missing values which is redundant as they share the same information.
ID's will consequently be dropped from the data, leaving 11 remaining variables 
in the data frame.
```{r Drop IDs}
df_trips <- df_trips %>% dplyr::select(-start_station_id, -end_station_id)
```

### Data Wrangling
Again, the purpose of this report is to find meaningful insights by 
profiling different customers and informing the marketing team of said new insights.
It may pertain to customer preferences, marketing efforts, ad placements etc. 
Hence there is a need to transform parts of the data to make it more suitable
for analysis.

#### Trip length
Since the gathered timestamps consist of both start and end times of each trip, 
calculating the trip length is possible by either subtracting one column
by another, or for fun try out an overly complicated function and change the data type 
three times over.
```{r Trip length}
# gather time difference, convert to seconds and then to hour-minutes-seconds format.
df_trips$trip_length <- df_trips %>% with(hms::hms(seconds_to_period(difftime(ended_at,started_at,units = "secs"))))
df_trips$trip_length %>%  sample(5)
df_trips$trip_length %>% min()
```
There are negative values that will be kept but filtered out for later analysis.
By doing so, there wont be a loss off valuable observations for other variables, 
like rideable type.

#### Distance travelled
Following code is an attempt at trying to make use of the recorded spatial coordinates of each observation. 
It will likely be inaccurate, as Haversine formula calculates the great-circle 
distance between two coordinates with no regards to different routes, while also
returning values of zero for people returning to the same starting station. 
But it may still give an idea of any differences between types of riders as 
there are close to 5.5 million observations.
```{r Distance travelled in meters}
df_trips <- df_trips %>% mutate(distance_travelled = distHaversine(cbind(start_lat, start_lng), cbind(end_lat, end_lng)))
df_trips$distance_travelled <- df_trips$distance_travelled %>% round(digits=2)
sample(df_trips$distance_travelled, 5)
```
#### Time variables
Divvying up the date-time column may reveal any possible trends in traffic, 
both by season, weekdays, and hour of day. To do so, assigning values to each 
month and weekday helps to plot the values in order along the x-axis.
```{r Time variables}
df_trips$month <- df_trips$started_at %>% strftime('%B')
df_trips$month <- ordered(df_trips$month, levels=c("January", "February", "March", "April", 
"May", "June", "July", "August", "September", "October", "November", "December"))
sample(df_trips$month,5)

df_trips$weekday <- df_trips$started_at %>% strftime('%A')
df_trips$weekday <- ordered(df_trips$weekday, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", 
"Friday", "Saturday", "Sunday"))
sample(df_trips$weekday,5)

df_trips$hour <- df_trips$started_at %>% strftime('%H')
sample(df_trips$hour,5)
```

#### Data overview
One final look before proceeding with the analysis.
```{r Data summary}
df_trips %>%
  glimpse()
```
Now there are 16 variables to work with, each to be applied and explored
for the benefit of the marketing team. Note again that distance
traveled is highly inaccurate but making use of every piece of data could be 
considered good practice. It might also make a good reference for future cases.

### Analysis
With the data cleaned, adjusted and prepared the analysis can finally begin. It will 
cover ratios, time series, and frequencies for gaining insights into customer profiles. 
This would help adjust marketing efforts, improve engagement, and better promote Cyclistic
with the casual user base. 

```{r Proportions, class.source = 'fold-hide'}
plot_prop <- df_trips %>% 
  count(member_casual) %>% 
  mutate(pct=prop.table(n)) %>%
  ggplot(aes("", pct, fill = member_casual)) + 
  geom_bar(width = 1, size = 1, color = "white", stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = scales::percent(pct)),
            position = position_stack(vjust = 0.5)) +
  theme_minimal() +
  theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid  = element_blank(),
          legend.position = "bottom") +
  labs(x = NULL, y = NULL, title="Percantage of Members & Casual users")

plot_bikes <- df_trips %>% count(member_casual, rideable_type) %>% 
  mutate(pct=prop.table(n)) %>% 
  ggplot(aes(rideable_type, pct, fill = member_casual, 
             label = scales::percent(pct))) + 
  geom_col(position="dodge") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(position = position_dodge(.9), 
            vjust = -0.5, 
            size = 3) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title="Percentage of rideable type", x="", face = "bold", size=14)

ggarrange(plot_prop, plot_bikes, 
          ncol = 2, 
          common.legend = TRUE,
          legend="bottom")
```

For an overview of the customer base, members account for 54.8% of total rides 
taken during the 1-year period, while casual riders comprised the remaining 
45.2% of trips. There is clearly a strong incentive for Cyclistic to convert 
casual riders into members, rather than attracting new consumers. 
Besides the fact that classic bikes are the most popular option for both groups,
neither favor the docked bikes. 

```{r Time series, message=FALSE, warning=FALSE, class.source = 'fold-hide'}
plot_month <- df_trips %>% ggplot(aes(month, color=member_casual, group = member_casual)) + 
  geom_line(stat='count', size = 1) +
  geom_point(stat='count', size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 7.7),
        legend.position = "none")

plot_weekday <- df_trips %>% ggplot(aes(weekday, color=member_casual, group = member_casual)) + 
  geom_line(stat='count', size = 1) +
  geom_point(stat='count', size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 7.7),
        legend.position = "none") +
  scale_y_continuous(labels = scientific)

plot_hour <- df_trips %>% ggplot(aes(hour, color=member_casual, group = member_casual)) + 
  geom_line(stat='count', size = 1) +
  geom_point(stat='count', size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 7.7),
        legend.position = "none")
  
legend <- df_trips %>% ggplot(aes(month, color=member_casual, group = member_casual)) + 
  geom_line(stat='count', size = 1) +
  geom_point(stat='count', size = 2) +
  theme_minimal()
plot_legend <- get_legend(legend)

ggarrange(plot_month, plot_weekday, plot_hour, plot_legend, 
         align="hv") %>%
  annotate_figure(top = text_grob("Trips across 12 months", size=14))

```

In terms of monthly trips, number of casual users peak in July and hit a bottom
in February, a clear relationship to weather conditions and vacation times. 
Members do share similar numbers but, perhaps not unexpectedly, do have a flatter 
slope towards the end of the year since they are more likely to brave the cold
later into the year than what casual riders would.

In the line chart for weekdays, members are shown to have 
consistently high numbers throughout work days but experience dip during the weekends. 
Again in line with behaviour of workers. Even clearer when comparing to casual riders
showing considerably smaller numbers during weekdays but peaking on the weekends.  

Looking at traffic by hour of the day, members show clear spikes during rush hours
while casual users have overall lower numbers throughout the days. This is 
probably because it only accounts for starting hours and not following hours
within one trip, and casual riders make up a smaller portion against members.
The plot might have told a different story if it accounted for trip length, 
with casual riders likely taking longer trips.  

All three plots clearly reflects their status as casual riders and 
members, the first group rides for leisure and the latter rides for commuting.

```{r trip_distance_hour, class.source = 'fold-hide'}
plot_duration <- df_trips %>% filter(trip_length > 0) %>% 
  ggplot(aes(hour, trip_length, color=member_casual, group = member_casual)) + 
  stat_summary(fun = mean, geom = "line", size=1) +
  stat_summary(fun = mean, geom = "point", size=2) +# The stat_summary function removes all rows with null values.
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 7.7)) +
  theme_minimal() +
  ylab("trip length [h:m:s]")
plot_distance <- df_trips %>% filter(distance_travelled > 0) %>% 
  ggplot(aes(hour, distance_travelled, group = member_casual, color = member_casual)) +
  stat_summary(fun = mean, geom = "line", size=1) +
  stat_summary(fun = mean, geom = "point", size=2) + # The stat_summary function removes all rows with null values.
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 7.7)) +
  theme_minimal() +
  ylab("distance travelled [m]")

ggarrange(plot_duration, plot_distance, 
          common.legend = T, 
          legend = "bottom", 
          align = "hv", 
          ncol = 1) %>% 
  annotate_figure(top = text_grob("Average Trip length- and Distance travelled, by hour", 
                                  size=14))

```

Casual users on average tend to bike for both longer periods of time and 
longer distances, which is reasonable as workers would likely stick to a
few selected routes for making good on time. But also, casual users would likely
make stops and take sporadic routes for sightseeing. The overall distance is 
rather short, suggesting that they do not bike for sport or fitness. 

Do note that the visualization does not 
account for consumers that have presumably returned to the same starting station,
since those are filtered out to not drastically lower the average distance 
traveled. Those who were filtered out amounts to

```{r Controlling for distance_travelled & trip_length}
df_trips_distance_zero <- filter(df_trips, distance_travelled==0)
table(df_trips_distance_zero$member_casual)
```

which is close to 10% of total observations. 

```{r 5 common, class.source = 'fold-hide'}
df_trips %>% filter(trip_length > 0) %>%
  drop_na(start_station_name) %>% 
  add_count(start_station_name) %>%
  filter(dense_rank(-n) < 6) %>% 
  ggplot(aes(start_station_name, fill=member_casual, group = member_casual)) + 
  geom_bar(position=position_dodge()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        plot.title = element_text(hjust = 0.5)) +
  labs(title="5 most common stations", 
       x = "station name",
       y = "count")

```

'Streeter Dr & Grand Ave' and Millenium Park would be the best stations for 
placing advertisement appealing to casual riders. Members have very low numbers
at Millennium Park, likely because it is not en route to any workplaces. 
```{r Duration by common stations, class.source = 'fold-hide'}
df_trips %>% filter(trip_length > 0) %>%
  drop_na(start_station_name) %>% 
  add_count(start_station_name) %>%
  filter(dense_rank(-n) < 6) %>% 
  ggplot(aes(start_station_name, trip_length, fill=member_casual, group = member_casual)) + 
  stat_summary(fun = mean, geom = "bar", position=position_dodge()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        plot.title = element_text(hjust = 0.5)) +
  labs(title="Trip length by 5 most common stations", 
       y = "trip length [h:m:s]",
       x = "station name")
```

Trip length by the 5 most common starting stations support the above hypothesis,
as casual riders would likely make stops and have extended outings at Millennium Park.

```{r Common route, class.source = 'fold-hide'}
df_trips %>% drop_na(start_station_name) %>% 
  group_by(member_casual) %>% 
  drop_na(start_station_name) %>%
  drop_na(end_station_name) %>% 
  count(paste(start_station_name," - ", end_station_name), sort = TRUE)
```
The above output contain a list of routes by the number of counts. However,
not a lot of stock can be put into advertisement along these supposed routes since 
casual riders tend to have far longer trip length but pretty similar levels
to distance traveled in relation to members. Likely because they don't ride straight from 
station A to B and instead are sightseeing, taking breaks or following 
sporadic routes. They are also more inclined to return to the same starting station,
adding more value for advertising at the stations rather than along the routes. 

```{r Common m routes, class.source = 'fold-hide'}
df_trips %>% drop_na(start_station_name) %>% 
  filter(member_casual=="member") %>% 
  drop_na(start_station_name) %>%
  drop_na(end_station_name) %>% 
  count(paste(start_station_name," - ", end_station_name), sort = TRUE)
```
Members are less inclined to return to the same starting station, and would 
natural end at another station between work and home. Advertisement along common routes 
would have a far more effective reach on members than on casual riders, 
if that is an option Cyclistic would like to pursue. 

### Recommendations

* While a small portion of casual riders do favor the docked bikes, the majority
prefer classical and electrical bikes. Any promotion appealing to said 
demographic should include those selected bikes. 
* Best campaign period is between June and September.
* Any marketing should be placed around 'Streeter Dr & Grand Ave' and Millennium Park for 
most impressions. 
* Casual users take longer trips in the morning and are the most active in the 
evenings. Any active, on-the-ground advertisement from noon until evening would likely 
have a good reach. Naturally on the weekends.  
* Cyclistic could either promote commuting to work by bike, as that gives them 
an incentive to pay for a yearly membership, or launch a program that caters to 
their long ride tendencies through some sort of point system.

