---
title: "Data With Danny: Pizza Runner"
author: "Benjamin Sivac"
date: "`r Sys.Date()`"
output: 
  rmarkdown::github_document
---
```{r, include=FALSE}
hook_output <- knitr::knit_hooks$get("output")
```

```{r Loading packages, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(odbc)
library(tidyverse)
library(datamodelr) # ER diagram
```

```{r, include=FALSE}
con <- DBI::dbConnect(odbc::odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "dwd#2", 
                      Trusted_Connection = "True")
```

### Introduction
Did you know that over 115 million kilograms of pizza is consumed daily worldwide??? (Well according to Wikipedia anyway…)

Danny was scrolling through his Instagram feed when something really caught his eye - “80s Retro Styling and Pizza Is The Future!”

Danny was sold on the idea, but he knew that pizza alone was not going to help him get seed funding to expand his new Pizza Empire - so he had one more genius idea to combine with it - he was going to Uberize it - and so Pizza Runner was launched!

Danny started by recruiting “runners” to deliver fresh pizza from Pizza Runner Headquarters (otherwise known as Danny’s house) and also maxed out his credit card to pay freelance developers to build a mobile app to accept orders from customers.


### Available Data
Because Danny had a few years of experience as a data scientist - he was very aware that data collection was going to be critical for his business’ growth.

He has prepared for us an entity relationship diagram of his database design but requires further assistance to clean his data and apply some basic calculations so he can better direct his runners and optimise Pizza Runner’s operations.

All datasets exist within the pizza_runner database schema - be sure to include this reference within your SQL scripts as you start exploring the data and answering the case study questions.


#### Entity Relationship Diagram


#### Table 1: runners
The runners table shows the registration_date for each new runner

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  runners
```


#### Table 2: customer_orders
Customer pizza orders are captured in the customer_orders table with 1 row for each individual pizza that is part of the order.


```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  customer_orders
```

The pizza_id relates to the type of pizza which was ordered whilst the exclusions are the ingredient_id values which should be removed from the pizza and the extras are the ingredient_id values which need to be added to the pizza.

Note that customers can order multiple pizzas in a single order with varying exclusions and extras values even if the pizza is the same type!

The exclusions and extras columns will need to be cleaned up before using them in your queries.



#### Table 3: runner_orders
After each orders are received through the system - they are assigned to a runner - however not all orders are fully completed and can be cancelled by the restaurant or the customer.


```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  runner_orders
```

The pickup_time is the timestamp at which the runner arrives at the Pizza Runner headquarters to pick up the freshly cooked pizzas. The distance and duration fields are related to how far and long the runner had to travel to deliver the order to the respective customer.

There are some known data issues with this table so be careful when using this in your queries - make sure to check the data types for each column in the schema SQL!


#### Table 4: pizza_names
At the moment - Pizza Runner only has 2 pizzas available the Meat Lovers or Vegetarian!

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  pizza_names
```

#### Table 5: pizza_recipes
Each pizza_id has a standard set of toppings which are used as part of the pizza recipe.


```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  pizza_recipes
```

#### Table 6: pizza_toppings
This table contains all of the topping_name values with their corresponding topping_id value

```{sql echo=FALSE, connection=con}
SELECT
  *
FROM 
  pizza_toppings
```


### Case Study Questions
This case study has LOTS of questions - they are broken up by area of focus including:

* Pizza Metrics
* Runner and Customer Experience
* Ingredient Optimisation
* Pricing and Ratings

Each of the following case study questions can be answered using a single SQL statement.

Again, there are many questions in this case study - please feel free to pick and choose which ones you’d like to try!


#### Clean tables
Before you start writing your SQL queries however - you might want to investigate the data, you may want to do something with some of those null values and data types in the customer_orders and runner_orders tables!

Clean customer_orders table:
```{sql, connection=con}
UPDATE dbo.customer_orders 
SET exclusions = NULLIF(exclusions, 'null'),
	extras = NULLIF(extras, 'null');

UPDATE dbo.customer_orders 
SET 
	exclusions = NULLIF(exclusions, ''),
	extras = NULLIF(extras, '');

-- See end result
SELECT
  *
FROM
  customer_orders
```

```{sql echo=FALSE, connection=con}
-- See end result:
SELECT
  *
FROM
  customer_orders
```

Clean runner_orders table:
```{sql, connection=con}
-- Fix duration column.
UPDATE dbo.runner_orders 
SET 
	duration = TRIM(REPLACE(duration, 'minutes', ''))
UPDATE dbo.runner_orders 
SET 
	duration = TRIM(REPLACE(duration, 'mins', ''));
UPDATE dbo.runner_orders 
SET 
	duration = TRIM(REPLACE(duration, 'null', ''));
UPDATE dbo.runner_orders 
SET 
	duration = TRIM(REPLACE(duration, 'minute', ''));

ALTER TABLE dbo.runner_orders 
ALTER COLUMN duration FLOAT;

-- Fix distance column
UPDATE dbo.runner_orders 
SET 
	distance = TRIM(REPLACE(distance, 'km', ''))
UPDATE dbo.runner_orders 
SET 
	distance = TRIM(REPLACE(distance, 'null', ''));
ALTER TABLE dbo.runner_orders 
ALTER COLUMN distance FLOAT;

-- Fix cancellation column
UPDATE dbo.runner_orders 
SET 
	cancellation = ISNULL(cancellation, '');
UPDATE dbo.runner_orders 
SET 
	cancellation = TRIM(REPLACE(cancellation, 'null', ''));

-- Fix pickup_time column
UPDATE dbo.runner_orders 
SET 
	pickup_time = NULLIF(pickup_time, 'null');
ALTER TABLE dbo.runner_orders 
ALTER COLUMN pickup_time DATETIME2;

ALTER TABLE dbo.pizza_names
ALTER COLUMN pizza_name NVARCHAR(100);

-- See end result:
SELECT
  *
FROM
  runner_orders
```

```{sql echo=FALSE, connection=con}
-- See end result:
SELECT
  *
FROM
  runner_orders
```

***

#### A. Pizza Metrics

**1. How many pizzas were ordered?**
```{sql, connection = con}
SELECT
	COUNT(*) AS ordered_pizzas
FROM
	customer_orders;
```
***

**2. How many unique customer orders were made?**
```{sql, connection = con}
SELECT
	customer_id,
	COUNT(customer_id) AS nbr_of_orders
FROM
	customer_orders
GROUP BY customer_id;
```
***

**3. How many successful orders were delivered by each runner?**
```{sql, connection = con}
SELECT
	runner_id,
	COUNT(order_id) AS succesful_orders
FROM
	runner_orders
WHERE pickup_time IS NOT NULL
GROUP BY runner_id;
```
***

**4. How many of each type of pizza was delivered?**
```{sql, connection = con}
SELECT
	pizza_id,
	COUNT(pizza_id) AS delievered
FROM
	runner_orders ro
JOIN
	customer_orders co
	ON ro.order_id = co.order_id
WHERE pickup_time IS NOT NULL
GROUP BY pizza_id;
```
***

**5. How many Vegetarian and Meatlovers were ordered by each customer?**
```{sql, connection = con}
SELECT
	customer_id,
	pizza_name,
	COUNT(order_id) AS orders
FROM
	customer_orders co
JOIN
	pizza_names pn
	ON co.pizza_id=pn.pizza_id
GROUP BY co.customer_id, pizza_name
ORDER BY customer_id
```
***

**6. What was the maximum number of pizzas delivered in a single order?**
```{sql, connection = con}
WITH cte_pizza_count AS (
	SELECT 
			COUNT(co.order_id) as orders 
		FROM customer_orders co 
		JOIN runner_orders ro 
			ON co.order_id=ro.order_id 
		WHERE pickup_time IS NOT NULL 
		GROUP BY co.order_id
)

SELECT
	MAX(orders) AS most_pizzas_ordered
FROM
	cte_pizza_count;
```
***

**7. For each customer, how many delivered pizzas had at least 1 change and how many had no changes?**
```{sql, connection = con}
SELECT
	customer_id,
	SUM(CASE WHEN exclusions IS NOT NULL OR extras IS NOT NULL THEN 1 ELSE 0 END) AS orders_with_changes,
	SUM(CASE WHEN exclusions IS NULL AND extras IS NULL THEN 1 ELSE 0 END) AS orders_without_changes
FROM 
	customer_orders co 
JOIN 
	runner_orders ro 
	ON co.order_id=ro.order_id 
WHERE pickup_time IS NOT NULL 
GROUP BY customer_id;
```
***

**8. How many pizzas were delivered that had both exclusions and extras?**
```{sql, connection = con}
SELECT
	SUM(CASE WHEN exclusions IS NOT NULL AND extras IS NOT NULL THEN 1 ELSE 0 END) AS excluded_and_added
FROM 
	customer_orders co 
JOIN 
	runner_orders ro 
	ON co.order_id=ro.order_id 
WHERE pickup_time IS NOT NULL;
```
***

**9. What was the total volume of pizzas ordered for each hour of the day?**
```{sql, connection = con}
SELECT
	DATEPART(HOUR,order_time) AS hour_of_day,
	COUNT(order_id) AS nbr_pizza_ordered
FROM
	customer_orders
GROUP BY DATEPART(HOUR,order_time);
```
***

**10. What was the volume of orders for each day of the week?**
```{sql, connection = con}
SELECT
	DATENAME(WEEKDAY,order_time) AS day_of_week,
	COUNT(order_id) AS nbr_pizza_ordered
FROM
	customer_orders
GROUP BY DATENAME(WEEKDAY,order_time);
```
***

#### B. Runner and Customer Experience

**1. How many runners signed up for each 1 week period? (i.e. week starts 2021-01-01)**
```{sql, connection = con}
SELECT
	DATEPART(WEEK,registration_date) AS week,
	COUNT(runner_id) AS nbr_runners
FROM
	runners
GROUP BY DATEPART(WEEK,registration_date);
```
***

**2. What was the average time in minutes it took for each runner to arrive at the Pizza Runner HQ to pickup the order?**
```{sql, connection = con}
SELECT
	AVG(duration) AS avg_pickup_time_mins
FROM
	runner_orders;
```
***

**3. Is there any relationship between the number of pizzas and how long the order takes to prepare?**
```{sql, connection = con}
WITH cte_prep_time AS(
	SELECT
		COUNT(co.order_id) AS number_of_pizzas,
		AVG(DATEDIFF(MINUTE, order_time, pickup_time)) AS prep_time
	FROM
		runner_orders ro
	JOIN
		customer_orders co
		ON	ro.order_id = co.order_id
	WHERE pickup_time IS NOT NULL
	GROUP BY co.order_id
)
SELECT
	number_of_pizzas,
	AVG(prep_time)
FROM
	cte_prep_time
GROUP BY number_of_pizzas;
```
***

**1. How many runners signed up for each 1 week period? (i.e. week starts 2021-01-01)**
```{sql, connection = con}
SELECT
	DATEPART(WEEK,registration_date) AS week,
	COUNT(runner_id) AS nbr_runners
FROM
	runners
GROUP BY DATEPART(WEEK,registration_date);
```
***

**4. What was the average distance travelled for each customer?**
```{sql, connection = con}
SELECT
	customer_id,
	ROUND(AVG(distance),2) AS avg_distance
FROM
	customer_orders co
JOIN
	runner_orders ro
	ON	co.order_id=ro.order_id
WHERE pickup_time IS NOT NULL
GROUP BY customer_id;
```
***

**5. What was the difference between the longest and shortest delivery times for all orders?**
```{sql, connection = con}
SELECT
	MAX(DATEDIFF(MINUTE, order_time, pickup_time)) AS max_delivery_difference
FROM
	runner_orders ro
JOIN
	customer_orders co
	ON	ro.order_id = co.order_id
WHERE pickup_time IS NOT NULL;
```
***

**6. What was the average speed for each runner for each delivery and do you notice any trend for these values?**
```{sql, connection = con}
SELECT
	runner_id,
	co.order_id,
	COUNT(co.order_id) AS nbr_pizzas,
	distance,
	duration,
	ROUND(AVG(distance/duration*60),2) AS km_h
FROM
	runner_orders ro
JOIN
	customer_orders co
	ON	ro.order_id = co.order_id
WHERE duration != 0
GROUP BY runner_id, co.order_id, distance, duration
ORDER BY runner_id, co.order_id;
```
***

**7. What is the successful delivery percentage for each runner?**
```{sql, connection = con}
SELECT
	runner_id,
	FORMAT(CAST(SUM(CASE WHEN pickup_time IS NOT NULL THEN 1 ELSE 0 END) AS FLOAT) / COUNT(*), 'p') AS delivery_pct
FROM
	runner_orders
GROUP BY runner_id;
```
***

#### C. Ingredient Optimisation

**1. What are the standard ingredients for each pizza?**
```{sql, connection = con}
WITH cte_split AS (
	SELECT
		*
	FROM
		pizza_recipes CROSS APPLY STRING_SPLIT(toppings,',')
)

SELECT
	pn.pizza_name,
	STRING_AGG(CAST(topping_name AS VARCHAR), ', ') AS ingredients
FROM
	cte_split t
JOIN
	dbo.pizza_toppings pt
	ON	t.value = pt.topping_id
JOIN
	pizza_names pn
	ON	t.pizza_id = pn.pizza_id
GROUP BY pn.pizza_name;
```
***

**2. What was the most commonly added extra?**
```{sql, connection = con, eval=F, echo=T}
WITH cte_split AS(
	SELECT
		value AS topping,
		COUNT(*) AS count
	FROM
		customer_orders co CROSS APPLY STRING_SPLIT(extras,',')
	GROUP BY value
)
SELECT TOP 1
	topping_name AS most_common_extra,
	count
FROM
	cte_split t
JOIN
	pizza_toppings pt
	ON	t.topping = pt.topping_id
ORDER BY count DESC;
```
This and the following code ran properly on SQL server but I kept getting "Invalid Descriptor Index" trying to run it in rmarkdown... I'll just add pictures of the result from SQL server. 

***

**3. What was the most common exclusion?**
```{sql, connection = con, eval=F, echo=T}
WITH cte_split AS(
	SELECT
		value AS topping,
		COUNT(*) AS count
	FROM
		customer_orders co CROSS APPLY STRING_SPLIT(exclusions,',')
	GROUP BY value
)
SELECT TOP 1
	topping_name AS most_excluded_topping,
	count
FROM
	cte_split t
JOIN
	pizza_toppings pt
	ON	t.topping = pt.topping_id
ORDER BY COUNT DESC;
```
***

**4. Generate an order item for each record in the customers_orders table in the format of one of the following:**

* **Meat Lovers**
* **Meat Lovers - Exclude Beef**
* **Meat Lovers - Extra Bacon**
* **Meat Lovers - Exclude Cheese, Bacon - Extra Mushroom, Peppers**
```{sql, connection = con}
SELECT
	order_id,
	CASE
		WHEN CHARINDEX('1', extras) > 0 THEN 'Meat Lovers - Extra Bacon'
		WHEN CHARINDEX('3', exclusions) > 0 THEN 'Meat Lovers - Exclude Beef'
		WHEN CHARINDEX('4', exclusions) > 0 AND CHARINDEX('1', exclusions) > 0
			AND CHARINDEX('1', extras) > 0 AND CHARINDEX('6', extras) > 0 
			AND CHARINDEX('9', extras) > 0 THEN 'Meat Lovers - Exclude Cheese, Bacon - Extra Mushroom, Peppers'
		WHEN pizza_id=1 THEN 'Meat Lovers' 
		ELSE 'Vegetarian' END as ordered_item
FROM
	customer_orders
```
Not a very dynamic or elegant solution, but the question had very specific instructions. I did however add 'Vegetarian' as one label.

***

#### D. Pricing and Ratings

**1. If a Meat Lovers pizza costs $12 and Vegetarian costs $10 and there were no charges for changes - how much money has Pizza Runner made so far if there are no delivery fees?**
```{sql, connection = con}
SELECT
	SUM(CASE WHEN pizza_id=1 THEN 12 ELSE 10 END) AS total_income
FROM
	customer_orders co
JOIN
	runner_orders ro
	ON	co.order_id=ro.order_id
WHERE pickup_time IS NOT NULL;
```
***

**2. What if there was an additional $1 charge for any pizza extras?**
```{sql, connection = con}
SELECT
	SUM(CASE 
			WHEN pizza_id=1 AND extras IS NOT NULL THEN 13
			WHEN pizza_id=1 THEN 12
			WHEN pizza_id=2 AND extras IS NOT NULL THEN 11 ELSE 10 END) AS total_income
FROM
	customer_orders co
JOIN
	runner_orders ro
	ON	co.order_id=ro.order_id
WHERE pickup_time IS NOT NULL;
```
***





















